---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PCAtest

<!-- badges: start -->
<!-- badges: end -->

The goal of PCAtest is to evaluate the overall significance of a PCA, of each PC
axis, and of the contributions of each observed variable to the significant axes 
based on permutation-based statistical tests. PCAtest uses random permutations 
to build null distributions for several statistics of a PCAanalysis: Psi (Vieira
2012),Phi (Gleason and Staelin 1975), the rank-of-roots (ter Braak 1988), the 
index of the loadings (Vieira 2012), and the correlations of the PC with the 
variables (Jackson 1991). Comparing these distributions with the observed values
of the statistics, the function tests: (1) the hypothesis that there is more 
correlational structure among the observed variables than expected by random 
chance, (2) the statistical significance of each PC, and (3) the contribution of
each observed variable to each significant PC. The function also calculates the 
sampling variance around mean observed statistics based on bootstrap replicates.

## Installation

You can install the released and the development versions from [GitHub]
(https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("arleyc/PCAtest")
```
## Examples

```{r example1}
library(PCAtest)
# PCA analysis of five uncorrelated variables
data("ex0")
result<-PCAtest(ex0, 100, 100, 0.05, varcorr=FALSE, counter=FALSE, plot=TRUE)
```

```{r example2}
#PCA analysis of five correlated (r=0.5) variables
data("ex05")
result<-PCAtest(ex05, 100, 100, 0.05, varcorr=FALSE, counter=FALSE, plot=TRUE)
```

```{r example3}
#PCA analysis of five correlated (r=1) variables
v1<-seq(0,1,0.01)
v5=v4=v3=v2=v1
ex1<-cbind(v1,v2,v3,v4,v5)
result<-PCAtest(ex1, 100, 100, 0.05, varcorr=FALSE, counter=FALSE, plot=TRUE)
```

```{r example4}
#PCA analysis of seven morphological variables from 29 ant species (from
# Wong and Carmona 2021,  https://doi.org/10.1111/2041-210X.13568)
data("ants")
result<-PCAtest(ants, 100, 100, 0.05, varcorr=FALSE, counter=FALSE, plot=TRUE)
```

## Testing
[Automated testing with testthat](tests/testthat/testthat.md)
