---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PCAtest

<!-- badges: start -->
<!-- badges: end -->

The goal of PCAtest is to evaluate the overall significance of a PCA, of each PC
axis, and of the contributions of each observed variable to the significant axes 
based on permutation-based statistical tests. PCAtest uses random permutations to
build null distributions for several statistics of a PCAanalysis: Psi (Vieira 2012),
Phi (Gleason and Staelin 1975), the rank-of-roots (ter Braak 1988), the index of the
loadings (Vieira 2012), and the correlations of the PC with the variables (Jackson 
1991). Comparing these distributions with the observed values of the statistics, the 
function tests: (1) the hypothesis that there is more correlational structure among 
the observed variables than expected by random chance, (2) the statistical significance
of each PC, and (3) the contribution of each observed variable to each significant PC. 
The function also calculates the sampling variance around mean observed statistics based
on bootstrap replicates.

## Installation

You can install the released and the development versions from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("arleyc/PCAtest")
```
## Example

```{r example1}
library(PCAtest)
# PCA analysis of two uncorrelated variables
v1<-runif(100,0,1)
v2<-runif(100,0,1)
x<-cbind(v1,v2)
result<-PCAtest(x, 100, 100, 0.05, corr=FALSE, plot=TRUE)
```

```{r example2}
#PCA analysis of two fully-correlated variables
v1<-seq(0,1,0.01)
v2<-seq(0,1,0.01)
x<-cbind(v1,v2)
PCAout<-PCAtest(x, 100, 100, 0.05, corr=FALSE, plot=TRUE)
```
