---
title: "PCAtest: Statistical significance of PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PCAtest: Statistical significance of PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
Introduction
---
A PCA is meaningful and should only be applied if there is a significant correlation structure among at least some of the variables in a dataset. Due to sampling variance alone, it is possible that some level of random correlation structure could occur in the data. Therefore, it is necessary to evaluate if the observed correlation structure is higher than expected due to chance alone. 

PCAtest uses a pair of statistics, Psi and Phi, based on the eigenvalues (roots) of a PCA to evaluate its significance by comparing observed values with their null distributions. The expected values of Psi and Phi are calculated from random permutations of the observations within variables, which effectively removes any correlation structure among variables. If the observed statistics are not larger than the expected null distributions, PCA is meaningless and it should not be used.

If the null hypothesis of a non-significant PCA is rejected, then PCAtest calculates the observed eigenvalues and builds null distributions to select the number of significant PCs. In addition, PCAtest also calculates two other statistics: the index of the loadings and the correlations of each PC with the variables, to test the contribution of the observed variables to each significant PC.

Overall significance of a PCA
---
PCAtest should not reject the null hypothesis of a non-significant PCA, if there is no correlation structure among the variables. We can generate and analyze in PCAtest a set of uncorrelated variables:

```{r, out.width="80%", out.height="80%",fig.width = 500/72, fig.height = 500/72, fig.retina=2, dpi=100}
library(PCAtest)
data("ex0")
result<-PCAtest(ex0, 100, 100, 0.05, indload=FALSE, varcorr=FALSE, counter=FALSE, plot=TRUE)
```

Selecting the number of significant PCs
---
If a PCA is significant, PCAtest uses random permutations to test which PC axes are significant based on random permutation and calculation of the percentage of total variation explained by each PC axis. We can use the published dataset from Wong & Carmona (2021, https://doi.org/10.1111/2041-210X.13568) consisting of seven morphological variables measured in 29 ant species:

```{r, out.width="80%", out.height="80%",fig.width = 500/72, fig.height = 500/72, fig.retina=2, dpi=100}
data("ants")
result<-PCAtest(ants, 100, 100, 0.05, indload=FALSE, varcorr=FALSE, counter=FALSE, plot=TRUE)
```

Testing which variables have significant loadings
---
In addition to selecting the number of significant PC axes, we are usually interested in finding out which variables have significant loadings on each significant PC axis. PCAtest runs random permutations and plots index loadings for all significant PCs:

```{r, out.width="80%", out.height="80%",fig.width = 500/72, fig.height = 500/72, fig.retina=2, dpi=100}
data("ants")
result<-PCAtest(ants, 100, 100, 0.05, indload=TRUE, varcorr=FALSE, counter=FALSE, plot=TRUE)
```

Testing which variables have significant correlations
---
Instead of index loadings, we can choose to use variable correlations to evaluate their contributions to each significant PC axis:

```{r, out.width="80%", out.height="80%",fig.width = 500/72, fig.height = 500/72, fig.retina=2, dpi=100}
data("ants")
result<-PCAtest(ants, 100, 100, 0.05, indload=FALSE, varcorr=TRUE, counter=FALSE, plot=FALSE)
```
